---
title: "Overview_of_the_junctions_package"
author: "Thijs Janzen"
date: "5/6/2021"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Overview_of_the_junctions_package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width = 7, fig.height = 7, echo = TRUE)
library(junctions)
require(tidyr)
require(dplyr)
require(ggplot2)
require(magrittr)
```

# Overview of the junctions package

The junctions package has since it's creation in 2018 been updated with a range
of extended functionality, spawning many associated functions. This vignette is
aimed at providing a clear overview of the several functions available, and 
displaying how you can potentially use these functions.

## Simulation functions

Because many of the other functions rely on input data, we will first explore 
functions that simulate such data. These functions simulate the accumulation of 
junctions in a Wright-Fisher model, with random mating, fixed population size, 
non-overlapping generations and recombination modeled with the mean number of 
crossovers drawn from a Poisson distribution with the size in Morgan of the 
chromosome as the mean of the distribution. The location of the crossovers is 
drawn from a uniform distribution across the chromosome. In all simulations, 
only one pair of chromosomes per individual is simulated.

### Phased and unphased simulations
These are the most recent additions to the package, where the user can simulate
the process of junction accumulation, and the detection of ancestry at a given 
set of markers. These simulations are designed to reflect the type of data one 
might get upon sequencing individuals. Simulations aimed more at matching 
analytical results can be found later on. The data is returned in long data 
format, and the user can, if required, superimpose phasing inaccuracy or reduced
marker coverage. 

```{r phased unphased simulation}
simulated_pop <- sim_phased_unphased(pop_size = 1000,
                                     freq_ancestor_1 = 0.5,
                                     total_runtime = 100,
                                     size_in_morgan = 1,
                                     markers = sort(runif(n = 1000, 0, 1)),
                                     time_points = seq(0, 100, by = 10))
```

### Mathematical simulations
In Janzen et al. 2018, two separate scenario's were explored: either using a 
finite or an infinite chromosome, where finite or infinite was referring to the 
number of junctions possibly spaced on a chromosome. To retain the functionality
suggested in that paper, and also to retain the code used for that paper, this 
functionality is available in the functions \code{sim_fin_chrom} and 
\code{sim_inf_chrom}. These functions are aimed at performing stochastic 
simulations of the accumulation of junctions, and how the placing and spacing of 
markers affects the detection of such junctions. As such, the simulations return
the population average number of accumulated junctions, and the number of 
detected junctions (given markers, only for the infinite chromosome), but not 
the actual local ancestry.
```{r sim fin chrom}
junctions_fin_chrom <- sim_fin_chrom(pop_size = 1000,
                                     freq_ancestor_1 = 0.5,
                                     total_runtime = 200,
                                     morgan = 1,
                                     R = 100) # the number of crossover sites

junctions_inf_chrom <- sim_inf_chrom(pop_size = 1000,
                                     freq_ancestor_1 = 0.5,
                                     total_runtime = 200,
                                     morgan = 1,
                                     markers = 100)

plot(junctions_inf_chrom$avgJunctions,
     type = "l", lwd = 2,
     col = "darkgreen",
     xlab = "Time since admixture",
     ylab = "Number of junctions")
lines(junctions_fin_chrom$avgJunctions, col = "blue", lwd = 2)
lines(junctions_inf_chrom$detectedJunctions, col = "lightgreen", lwd = 2)
legend("topleft", legend = c("Finite chromosome", "Infinite chromosome",
                             "Infinite chromosome detected"),
       col = c("blue", "darkgreen", "lightgreen"), lty = 1, lwd = 2)
```

### Backcrossing
Lastly, one additional function is available, where instead of random mating, 
the individuals cross back with one of the parents, as used in 
Lavretsky et al. 2019. This function also focuses mainly on the average number
of junctions and not on the distribution of local ancestry
```{r backcrossing}
backcross_result <- sim_backcrossing(population_size = 1000,
                                     freq_ancestor_1 = 0.5,
                                     total_runtime = 20,
                                     size_in_morgan = 1,
                                     number_of_markers = 100)

plot(backcross_result$average_junctions,
     type = "l", lwd = 2,
     col = "darkgreen",
     xlab = "Time since admixture",
     ylab = "Number of junctions")
```

## Inferring the time since admixture

### Junction based methods
Across Janzen et al. 2018 and Janzen & Miró Pina 2022, several equations have 
been proposed to infer the time since admixture, depending on the data available. 
Firstly, following equation 14 in Janzen et al. 2018, one can calculate the time 
since admixture based on the number of junctions detected in one chromosome, 
given a number of regularly spaced markers:
```{r estimate_time}
estimate_time(J = 10, N = 1000, R = 1000, H_0 = 0.5, C = 1)
```

In reality, markers are rarely regularly spaced and the practical applications
of equation 14 of Janzen et al. 2018 are limited. Therefore, Janzen & Miró Pina 
propose an exact method (Equation 3) of calculating the expected number of 
junctions in a single chromosome, given the distribution of markers across the 
chromosome.
```{r estimate time markers}
estimate_time_one_chrom(J = 10, N = 1000, H_0 = 0.5,
                        marker_distribution = sort(runif(n = 1000, 0, 1)))
```

### Local ancestry based methods
Mathematically, the total number of junctions is very tractable, however this
discards all the ancestry data collected at the respective markers. Janzen & 
Miró Pina therefore provide an alternative likelihood based method, that also
allows for the joint inference of the time since admixture across data resulting
from multiple chromosomes from the same individual. It is assumed here that the
individual of interest is haploid, since only data from a single chromosome is 
available. 
For the example below, we make use of previously simulated data and pretend that
the different chromosomes resulting from different individuals are, in fact, 
sampled from the same individual. This is of course completely mathematically 
incorrect, but is used here to demonstrate how to use the functions.
The function requires input of a so called 'ancestry_matrix', which is a matrix
with three columns: 1) chromosome indicator, 2) position on respective
chromosome in Morgan and 3) local ancestry. The function returns the maximum 
likelihood estimate of the time, and the associated likelihood. 
```{r estimate time haploid}
ancestry_data <- subset(simulated_pop, simulated_pop$time == 100)
ancestry_matrix <- dplyr::select(ancestry_data, c(individual,
                                                  location, anc_chrom_1))
estimate_time_haploid(ancestry_matrix = ancestry_matrix,
                      N = 1000,
                      freq_ancestor_1 = 0.5)
```

If instead, you have diploid ancestry data, you can make use of the function
\code{estimate_time_diploid}. The function can calculate the ancestry assuming
either that the data is phased, or unphased.
Below, we take the same data, and add 'fake' individual information. We do this,
because the function \code{estimate_time_diploid} requires an input matrix with
the following five columns: 1) individual, 2) chromosome, 3) location along 
respective chromosome in Morgan, 4) ancestry chromosome 1, 5) ancestry 
chromosome 2. Please note that in the case of unphased data, it does not matter
how ancestry is split out across the two chromosomes. 
Note that although we are in possession of phasing information (because we are
using simulation output), we can 'fake' the data to be unphased, in which case
the analysis does not take into account phasing information. 
We can specify the analysis type, which focuses on how the likelihoods are 
grouped. By default, the likelihood is calculated jointly for all chromosomes 
within the same individual. Thus, if multiple individuals are provided, multiple
time estimates are returned. Instead, one can also specify to group the data
by chromosome, jointly optimizing across individuals (not recommended). Lastly,
one can specify to split out the data per individual chromosome and provide 
separate estimates for each chromosome found in each individual. 
```{r estimate time diploid}
ancestry_matrix <- dplyr::select(ancestry_data, c(individual,
                                                  location,
                                                  anc_chrom_1, anc_chrom_2))
ancestry_matrix <- cbind(rep(1, length(ancestry_matrix$individual)),
                         ancestry_matrix)

t_phased <- estimate_time_diploid(ancestry_information = ancestry_matrix,
                                  analysis_type = "individuals",
                                  phased = TRUE,
                                  pop_size = 1000,
                                  freq_ancestor_1 = 0.5)

t_unphased <- estimate_time_diploid(ancestry_information = ancestry_matrix,
                                    analysis_type = "individuals",
                                    phased = FALSE,
                                    pop_size = 1000,
                                    freq_ancestor_1 = 0.5)

t_phased
t_unphased
```
We thus find that in this specific case, the time estimate using unphased data 
is slightly more accurate that the phased case, although they are both close.

## Likelihood functions
If we would like to investigate the likelihood profile around the maximum, for 
instance to assess how accurate our estimate is, we can do so with the 
respective likelihood functions. 

```{r likelihood}
ancestry_matrix <- dplyr::select(ancestry_data, c(individual, location,
                                                  anc_chrom_1, anc_chrom_2))

time_points <- 80:120
ll_phased <- log_likelihood_diploid(ancestry_matrix,
                                    pop_size = 1000,
                                    freq_ancestor_1 = 0.5,
                                    t = time_points,
                                    phased = TRUE)
ll_unphased <- log_likelihood_diploid(ancestry_matrix,
                                      pop_size = 1000,
                                      freq_ancestor_1 = 0.5,
                                      t = time_points,
                                      phased = FALSE)

to_plot <- tibble::tibble(time_points, ll_phased, ll_unphased)
to_plot %>%
  tidyr::gather(key = "phasing", value = "loglikelihood", -time_points) %>%
  ggplot2::ggplot(
      ggplot2::aes(x = time_points, y = loglikelihood, col = phasing)) +
  ggplot2::geom_line()
```

Alternatively, we can calculate the haploid log likelihood, e.g. the likelihood
of observing the ancestry in one chromosome:
```{r likelihood haploid}
ancestry_matrix <- dplyr::select(ancestry_data, c(individual, location,
                                                  anc_chrom_1))

ll_haploid <- log_likelihood_haploid(ancestry_matrix,
                                     N = 1000,
                                     freq_ancestor_1 = 0.5,
                                     t = time_points)
plot(ll_haploid ~ time_points, type = "l",
     xlab = "Time since admixture",
     ylab = "Loglikelihood")
```

## Expected number of junctions
Mathematically, one can calculate the expected number of junctions. The first 
equation covered here covers the derivations obtained in Janzen et al. 2018:
```{r expected number of junctions}
number_of_junctions(N = 100, R = 100, H_0 = 0.5, C = 1, t = 1000)
```

This equation however, has to assume that the markers are equally spaced, that 
is, that the distance between each consecutive marker pair is equal. This is 
rarely the case in empirical data, and a better version is provided by the 
function \code{number_of_junctions_markers}:
```{r expected number of junctions markers}
number_of_junctions_markers(N = 100, H_0 = 0.5, t = 1000,
                            marker_distribution = sort(runif(100, 0, 1)))
```

Similarly, we can calculate the probability of observing a junction in between
two markers, given an amount of recombination (in Morgan) between these 
two markers:
```{r expected number of junctions di}
number_of_junctions_di(N = 100, H_0 = 0.5, t = 1000, di = 1e-5)
```

Lastly, we have included a function that calculates the expected number of 
junctions after a number of generations, in a backcrossing mating scheme. 

```{r expected number of junctions back crossing}
number_of_junctions_backcross(H_0 = 0.5, C = 1, t = 10)
```

## Mathematical expressions
Following Janzen et al. 2018, there are a number of equations that might be 
useful.
Firstly, there is an equation to calculate the upper limit of junctions that can
be packed in a chromosome, given the number of (regularly spaced) markers:
```{r K}
calc_k(N = 1000, R = 1000, H_0 = 0.5, C = 1)
```

Any time spent after the population has reached this upper limit remains 
"invisible" for inference purposes, and thus it can be useful to calculate the
upper limit of age inference, coined 'Maximum Accurate Time', following equation
15 in Janzen et al. 2018:
```{r MAT}
calculate_mat(N = 1000, R = 1000, H_0 = 0.5, C = 1)
```

Similarly, we can estimate the error induced in time inference, where the error
is given by the amount of time difference induced if we were to infer one 
junction less (or more), following Equations 3 & 4 in the Supplementary 
information of Janzen et al. 2018.
```{r error}
time_error(t = 30, N = 1000, R = 1000, H_0 = 0.5, C = 1)
```

This concludes a runthrough of all the available functions in the junctions 
package. Feel free to contact me if you feel anything is missing, or to create
an issue if you run into a problem. 
